---
title: "Final Project"
author: "Elysa, Kate and Pan"
date: "2016Äê10ÔÂ28ÈÕ"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Metropolis-Hastings

Beta(6,4) is the distribution we need to sample from. And the proposal function given is $$\phi_{prop}|\phi_{old}¡«Beta(c*\phi_{old},c*(1 - \phi_{old}))$$, which should be used in acceptance ratio. We also need to consider the fact that the distribution we are sampling from is not sysmetric.

There are four steps for our algorithm.

##c=1
Our code for the part with C=1 is shown as the following.
```{r}
set.seed(1)
#Defination of proposal function
proposalfunction <- function(c, phi_old){   
  rbeta(1, c*phi_old,c*(1-phi_old))
}
#The constant in proposal function
c<-1                                        
# MC algorithm
run_metropolis_MCMC <- function(c, iterations){   
  chain = array(dim = c(iterations+1,1))
  chain[1,] = runif(1)
  for (i in 1:iterations){
    proposal = proposalfunction(c,chain[i,])
    
    probab = dbeta(proposal,6,4)/dbeta(chain[i,1],6,4)*dbeta(chain[i,],c*proposal,c*(1-proposal))/dbeta(proposal,c*chain[i,],c*(1-chain[i,])) # Acceptance Ratio
    if (runif(1) < probab){
      chain[i+1,] = proposal
    }else{
      chain[i+1,] = chain[i,]
    }
  }
  return(chain)
}

chain=run_metropolis_MCMC(1, 10000)

chain_real <- c(10000,1)
for (i in 1:10000){
  chain_real[i]<- rbeta(1,6,4) #generation a real chain from beta (6,4)
}

par(mfrow=c(1,3))  #1 row, 3 columns
plot(chain,type = 'l');abline(h = mean(chain_real),col = 'red'); acf(chain); hist(chain,freq = FALSE)  #plot commands
curve(dbeta(x, 6, 4), add=TRUE)  #adding density function of beta(6,4) to histogram

```

As can be seen from the histogram figure, the histogram doesn't really match with the density function of beta(6,4).
```{r}
chain_real <- c(10000,1)
for (i in 1:10000){
  chain_real[i]<- rbeta(1,6,4) #generation a real chain from beta (6,4)
}

par(mfrow=c(1,3))  #1 row, 3 columns
plot(chain_real,type = 'l');abline(h = mean(chain_real),col = 'red'); acf(chain_real); hist(chain_real,freq = FALSE)  #plot commands
curve(dbeta(x, 6, 4), add=TRUE)  #adding density function of beta(6,4) to histogram

ks.test(chain, chain_real)
```

The figure is for the real distribution - namely the beta(6,4). The number given is for the Kolmogorov¨CSmirnov test. We can see the two sets of figure don't really match with each other. The p value we get from the K-S test is quite small, which means the result is bad. 

Then we try to use burn in and thining to improve the performance of the algorithm.We modify the code to the following.
```{r}
#Defination of proposal function
proposalfunction <- function(c, phi_old){   
  rbeta(1, c*phi_old,c*(1-phi_old))
}
#The constant in proposal function
c<-1                                        
# MC algorithm
run_metropolis_MCMC <- function(c, iterations){   
  chain = array(dim = c(iterations+1,1))
  chain[1,] = runif(1)
  for (i in 1:iterations){
    proposal = proposalfunction(c,chain[i,])
    probab = dbeta(proposal,6,4)/dbeta(chain[i,1],6,4)*dbeta(chain[i,],c*proposal,c*(1-proposal))/dbeta(proposal,c*chain[i,],c*(1-chain[i,])) # Acceptance Ratio, no min here.
    if (runif(1) < probab){
      chain[i+1,] = proposal
    }else{
      chain[i+1,] = chain[i,]
    }
  }
  return(chain)
}


chain=run_metropolis_MCMC(1, 10000)

burnIn = 1000

par(mfrow=c(1,3))  #1 row, 3 columns
plot(chain[-(1:burnIn),1],type = 'l');abline(h = mean(chain_real),col = 'red'); acf(chain[-(1:burnIn),1]);hist(chain[-(1:burnIn),1],freq = FALSE)  #plot commands
curve(dbeta(x, 6, 4), add=TRUE)  #adding density function of beta(6,4) to histogram


```

We can find that the probability density function is more close to the the real one. 

##C=0.1
```{r}
#Defination of proposal function
proposalfunction <- function(c, phi_old){   
  rbeta(1, c*phi_old,c*(1-phi_old))
}
#The constant in proposal function
c<-0.1                                        
# MC algorithm
run_metropolis_MCMC <- function(c, iterations){   
  chain = array(dim = c(iterations+1,1))
  chain[1,] = runif(1)
  for (i in 1:iterations){
    proposal = proposalfunction(c,chain[i,])
    
    probab = dbeta(proposal,6,4)/dbeta(chain[i,1],6,4)*dbeta(chain[i,],c*proposal,c*(1-proposal))/dbeta(proposal,c*chain[i,],c*(1-chain[i,])) # Acceptance Ratio
    if (runif(1) < probab){
      chain[i+1,] = proposal
    }else{
      chain[i+1,] = chain[i,]
    }
  }
  return(chain)
}

chain=run_metropolis_MCMC(1, 10000)

par(mfrow=c(1,3))  #1 row, 3 columns
plot(chain,type = 'l');abline(h = mean(chain_real),col = 'red'); acf(chain); hist(chain,freq = FALSE)  #plot commands
curve(dbeta(x, 6, 4), add=TRUE)  #adding density function of beta(6,4) to histogram
ks.test(chain, chain_real)
```

##c=2.5
```{r}
#Defination of proposal function
proposalfunction <- function(c, phi_old){   
  rbeta(1, c*phi_old,c*(1-phi_old))
}
#The constant in proposal function
c<-2.5                                        
# MC algorithm
run_metropolis_MCMC <- function(c, iterations){   
  chain = array(dim = c(iterations+1,1))
  chain[1,] = runif(1)
  for (i in 1:iterations){
    proposal = proposalfunction(c,chain[i,])
    
    probab = dbeta(proposal,6,4)/dbeta(chain[i,1],6,4)*dbeta(chain[i,],c*proposal,c*(1-proposal))/dbeta(proposal,c*chain[i,],c*(1-chain[i,])) # Acceptance Ratio
    if (runif(1) < probab){
      chain[i+1,] = proposal
    }else{
      chain[i+1,] = chain[i,]
    }
  }
  return(chain)
}

chain=run_metropolis_MCMC(1, 10000)

par(mfrow=c(1,3))  #1 row, 3 columns
plot(chain,type = 'l');abline(h = mean(chain_real),col = 'red'); acf(chain); hist(chain,freq = FALSE)  #plot commands
curve(dbeta(x, 6, 4), add=TRUE)  #adding density function of beta(6,4) to histogram
ks.test(chain, chain_real)
```

##c=10
```{r}
#Defination of proposal function
proposalfunction <- function(c, phi_old){   
  rbeta(1, c*phi_old,c*(1-phi_old))
}
#The constant in proposal function
c<-10                                        
# MC algorithm
run_metropolis_MCMC <- function(c, iterations){   
  chain = array(dim = c(iterations+1,1))
  chain[1,] = runif(1)
  for (i in 1:iterations){
    proposal = proposalfunction(c,chain[i,])
    
    probab = dbeta(proposal,6,4)/dbeta(chain[i,1],6,4)*dbeta(chain[i,],c*proposal,c*(1-proposal))/dbeta(proposal,c*chain[i,],c*(1-chain[i,])) # Acceptance Ratio
    if (runif(1) < probab){
      chain[i+1,] = proposal
    }else{
      chain[i+1,] = chain[i,]
    }
  }
  return(chain)
}

chain=run_metropolis_MCMC(1, 10000)

par(mfrow=c(1,3))  #1 row, 3 columns
plot(chain,type = 'l'); abline(h = mean(chain_real),col = 'red');acf(chain); hist(chain,freq = FALSE)  #plot commands
curve(dbeta(x, 6, 4), add=TRUE)  #adding density function of beta(6,4) to histogram
ks.test(chain, chain_real)
```

We think C=10 gives the best estimate. This can be proved by the fact that the p value we get from the K-S test is the biggest when c=10 as well as the fact that the histogram matches best with the density function of Beta(6,4) when c=10.

Sometimes, we found that c=2.5 gives us the best result. I think the reason for that is because the random variable is ramdom.
