---
title: "K means final project"
author: "Elysa Strunin, Ekaterina Mikhailova, Shenrui Pan."
date: '1 ноября 2016 г '
output: html_document
---
We developed K means algorithm wich includes further steps:

Step 1) Randomly assign center locations for each of k clusters

Step 2) Assign each data point to the cluster whose center is closest to that point

Step 3) Recalculate each cluster center as the mean location of all the points assigned to it.

Check to make sure each cluster has at least one point.

Repeat steps 2-3.
```{r}
data(wine, package="rattle")
#Remove column 1 from the analysis
wine <- wine[-1]

#Step 1. Randomly assign 3 epicenters
K_means<-function(x,k){ 
  epicenters <- matrix(0, nrow=3, ncol=13)
  #Sample from each column to obtain an epicenter coordinate
  set.seed(2016)
  for(i in 1:k) {
    for(j in 1:ncol(x)) {
      col_sample <- sample(x[,j], 1)
      epicenters[i,j] <- col_sample
    }
  }
  
#Step 2. Calculate euclidian distances, take the min of each row and assign that column as the cluster; put it in fourth column.
  all_norms <- matrix(0, nrow=nrow(x), ncol=k+2)
  for(i in 1:k) {
    for(j in 1:nrow(x)) {
      e_dist <- dist(rbind(x[j,], epicenters[i,]), method = "euclidean")
      all_norms[j, i] <- e_dist
    }
  }
  for(i in 1:nrow(x)) {
    min_dist_index <- which.min(as.vector(all_norms[i, c(1:k)]))
    all_norms[i,k+1] <- min_dist_index
  }  
#Taking count of iterations. If clusters on iteration t-1 and t are equal then break and print the output.
  count <- 0
  repeat { count <- count + 1
  if  (identical(all_norms[,k+1], all_norms[,k+2])==TRUE) 
  {output <- all_norms[,k+1]
  print(output);
  break
  }else{
#Step 3. Recalculate the epicenters as the means per cluster
    for(i in 1:k) {
      wines_per_cluster <- wine[all_norms[,k+1]==i, ]
      column_means_per_cluster <- t(colMeans(wines_per_cluster))
      epicenters[i, ] <- column_means_per_cluster
    }
#Repeating Step 2 and 3: calculating euclidian distances
    for(i in 1:k) {
      for(j in 1:nrow(x)) {
        e_dist <- dist(rbind(x[j,], epicenters[i,]), method = "euclidean")
        all_norms[j, i] <- e_dist
      }
    }

#Take the minimum of each row; assign that column as the cluster on itertaion t; put it in k+2 column
    for(i in 1:nrow(x)) {
      min_dist_index <- which.min(as.vector(all_norms[i, c(1:k)]))
      all_norms[i,k+2] <- min_dist_index
    }  
    
#Checking that each cluster has at least one point,if not - break
    for (i in 1:k){
    if ((length(which(all_norms[,k+2]==i)) > 0) == FALSE ) 
    {
#If one of the clusters has less than one point - print the t-1 cluster assignments, which should have at least one point per cluster
      output <- all_norms[,k+1]
      print(output); break
    } 
    }
#If clusters on iteration t+1 (column k+1) and t (column k+2) are not equal, move the most recent clusters over and loop again
    if ((identical(all_norms[,k+1], all_norms[,k+2]))==FALSE) 
    {
      all_norms[,k+1] <- all_norms[,k+2]
      all_norms[,k+2] <- NA
    } 
    
  }} 
}



K_means(wine,3)
```

We grouped wine data into three clusters. Now we need to visualize our clustering results. 
```{r}
plotcluster(wine, output)
```

